Mixture Modeling Paper Results

### 2017-03-11 PaperEx170308.Rmd
EOD 170310 -Both algorithm and paper are in good shape -Upcoming Algorithm tasks: Overhaul to 4 key functions 1) Fitting: fitloops and bestfit    -BIC plot 2) Selection: fitpick uncert and multcut    -rawdist    -rawuncert    -rawhist 3) Cutoff: standard and non-standard cuts    -cutuncert    -cutdist 4) Summarization/Exportation: outputdata and summarytable

```{r loadfunctions, echo=F, warning=F, message=F}
library(mixsmsn)
library(sn)
#-----------------------------------------------
#[IC] Load Functions
#-----------------------------------------------

@@

#Runs loops of mixture model over different numbers of components and the normal and skew-normal distribution. This function accepts a data frame of the data and identifiers (variables must be entitled data and id respectively). This is the longest function to run and may fail, but can be re-run with adjustments using various options based on the error message output when the function fails. The default values are to do both skewnormal and normal approximation, with up to 5 components and 10 loops per combination of distribution and number of components. It is not reccommended to change these unless the default function fails.
#accepts a dataframe with columns id and data
fitloops<-function(datawithids,loops=10,maxcp=5){
  dists="Both"
  datawithids<-unique(datawithids)
  data<-datawithids$data
  datapts<-length(data)
  norms<-vector("list")
  errors<-matrix(0, nrow=maxcp, ncol=2)
  loopErr = function(data,i,j,family) {
    tryCatch(smsn.mix(data, nu =3, g = i, get.init = TRUE, criteria = TRUE, group = TRUE, family = family, calc.im=FALSE, obs.prob=TRUE, iter.max=5000, kmeans.param=list(iter.max = 50, n.start = 5)),
             error = function(e) {message(paste("Error in loop", j, "of", family, "with", i, "components", sep=" ")); 
               NA}) 
  }

@@

      norms$mc[[i]][[j]]<-vector("list", 15)
      norms$ms[[i]][[j]]<-vector("list", 15)
      if (dists=="Both") {
        norms$mc[[i]][[j]]<-loopErr(data,i,j,"Normal")  
        norms$ms[[i]][[j]]<-loopErr(data,i,j,"Skew.normal")
      } else if (dists=="Normal"){ 
        norms$mc[[i]][[j]]<-loopErr(data,i,j,"Normal")
      } else if (dists=="Skew.normal"){
        norms$ms[[i]][[j]]<-loopErr(data,i,j,"Skew.normal")
      } else {
        stop(print("dists must be specified as 'Both' or 'Normal' or 'Skew.normal'"))
      }

@@

    fitres<-vector("list")
    for (i in 1:maxcp){
      fitres$no[[i]]<-vector("list", 15)
      fitres$sn[[i]]<-vector("list", 15)
      fitres$no[[i]]<-norms$mc[[i]][[nobj$lmin$mc[i]]]
      fitres$sn[[i]]<-norms$ms[[i]][[nobj$lmin$ms[i]]]
    }
    fitres$loops<-loops
    fitres$maxcp<-maxcp
    fitres$datawithids<-datawithids
    fitres$errors<-errors
    return(fitres)
  } else {
    if (which.max(errors)<=maxcp){
      dist<-"Normal"
    } else {
      dist<-"Skew.normal"
    }
    if (ic<=2){
      stop(paste("More than two loops of the mixture modeling algorithm failed for the", dist, "distribution with", ic, "component(s). First ensure that the necessary packages have been loaded correctly. If the packages are loaded correctly, check for extreme outliers (which may be dropped with extreme caution) or violations of skewnormal or normal distribution assumptions."))
    } else {
      stop(paste("More than two loops of the mixture modeling algorithm failed for the", dist, "distribution with", ic, "component(s). Consider choosing a lower number of components (the default is 5) via the maxcp option of the fitloops function."))
    }
  }  
}

@@

  if (nobj$dist=="no"){
    nobj$distp="Normal"
    nobj$distf="Normal"
  } else if (nobj$dist=="sn") {
    nobj$distp="Skew normal"
    nobj$distf="Skew.normal"
  } else {
    nobj$distp=""
  }

  model<-vector("list")
  model$sn2<-fitres$sn[[2]]
  model$snbest<-fitres$sn[[nobj$minind$sn[1]]]
  model$norm2<-fitres$no[[2]]
  model$normbest<-fitres$no[[nobj$minind$no[1]]]
  model$best<-nobj$best
  model$bictab<-data.frame(normal=c(NA)[rep(c(1), times=maxcp)], skew.normal=c(NA)[rep(c(1), times=maxcp)])
  model$bictab$normal<-nobj$no
  model$bictab$skew.normal<-nobj$sn
  componentnames<-vector()
  for (i in 1:maxcp){
    componentnames[i]<-paste(i, " Component(s):  ", sep="")  
  }    
  rownames(model$bictab)<-componentnames
  colnames(model$bictab)<-c("Normal", "Skew normal")
  model$summary<-data.frame(desc=c(paste(nobj$distp, "with", nobj$minind[nobj$dist][1,1], "components ", sep=" "), paste("Skew normal with", nobj$minind$sn, "components ", sep=" "), paste("Normal with", nobj$minind$no, "components ", sep=" "), "Skew normal with 2 components ", "Normal with 2 components "), bic=c(nobj$best$bic, nobj$sn[nobj$minind$sn], nobj$no[nobj$minind$no], nobj$sn[2], nobj$no[2]))
  rownames(model$summary)<-c("Best Overall", "Best Skew normal  ", "Best Normal", "Two Skew normal", "Two Normal")
  colnames(model$summary)<-c("Description         ", "BIC  ")

@@

bicgraph<-function(model,title="BIC by type and number of distributions",setcolor=c("magenta", "blue", "black")){
  color<-setcolor
  if (is.na(model$bictab$Normal[1])==F) {
    plot(model$bictab$Normal, pch=1, col=color[1], main=title, xlab="Number of distributions", ylab="BIC", type="o",  xaxt="n")
    lines(model$bictab$`Skew normal`, type="o", pch=2, lty=2, col=color[2])
  } else {
    plot(model$bictab$Normal, type="o", pch=2, lty=2, col=color[2], main=title, xlab="Number of distributions", ylab="BIC",  xaxt="n")
    lines(model$bictab$`Skew normal`, pch=1, col=color[1], type="o")    
  }  
  points(model$bestdesc$ncomp,model$best$bic,type="o", pch="O", col=color[3],cex=2)
  axis(1, at = seq(1, length(model$bictab$Normal), by = 1))
  suppressWarnings(legend("topright", col=c(color[1], color[2], color[3]), lty=c(1,2,0), pch = c(1, 2, 1), legend=c("Normal", "Skew normal", "Best by BIC"), cex=c(1,1,1), xjust=1, seg.len=3, title="Distribution"))

@@

  data=fitres$datawithids$data
  n=ncomp
  if (dist!="Normal" & dist!="Skew.normal") {
    stop("Distribution must be specified as either 'Normal' or 'Skew.normal'")
  } else if ((dist=="Normal" & is.na(fitres$no[[1]]$bic)==T)| (dist=="Skew.normal" & is.na(fitres$sn[[1]]$bic)==T) | ncomp>length(fitres$no) ){
    stop("Must choose a distribution and number of components that were fitted in the fitloops function")
  } else {
    if (dist=="Normal"){
      dist="no"
    } else {
      dist="sn"
    }

    if (dist=="no"){
      distp="Normal"
    } else if (dist=="sn") {
      distp="Skew Normal"
    } else {
      distp=""
    }

@@

@@

```{r example1}

examp1<-read.csv(file="/Users/sarahsullivan/Documents/SullivanDocs/Rwd/cutoff/ex1.csv")

hist(examp1$data, breaks=100, main="Histogram of Data")
dens<-density(examp1$data)
plot(dens, main="Density of Data")

fitloopsobject<-fitloops(datawithids=examp1)
bestfitsobject<-bestfits(fitres=fitloopsobject)
bicgraph(model=bestfitsobject)

fitpickobject<-fitpick(fitres=fitloopsobject,bestfits=bestfitsobject)

uncertobject<-uncert(fitpickobj=fitpickobject)
multcutobject<-multcut(uncertobj=uncertobject)

rawuncertgraph(uncertobj=uncertobject,multcutobj=multcutobject)
rawdistgraph(pickobj=fitpickobject)
rawhistcuts(uncertobj=uncertobject,multcutobj=multcutobject)

standindetobject<-standindet(uncertobj=uncertobject,multcutobj=multcutobject,cutcomp=1)
specindetobject<-specindet(uncertobj=uncertobject,multcutobj=multcutobject,cutcomp=1,certlevel=0.8)

cutuncertgraph(cutobj=standindetobject)
cutdistgraph(cutobj=standindetobject, pickobj=fitpickobject)
cutuncertgraph(cutobj=specindetobject)
cutdistgraph(cutobj=specindetobject, pickobj=fitpickobject)
outdataobject<-outputdata(standindet=standindetobject,specindet=specindetobject)
summaryobject<-summarytable(outdataobj=outdataobject, standindetobj=standindetobject, specindetobj=specindetobject)
outtable<-summaryobject$outtab
outtable

@@

```{r example2}

examp2<-read.csv(file="/Users/sarahsullivan/Documents/SullivanDocs/Rwd/cutoff/ex2.csv", colClasses=c("NULL", NA, NA))

hist(examp2$data, breaks=100, main="Histogram of Data")
dens<-density(examp2$data)
plot(dens, main="Density of Data")

#fitloopsobject<-fitloops(datawithids=examp2)
  #might not run due to 5 component solution not being easily optimized, if it doesn't run, either load the full file, or run fitloopsobject<-fitloops(datawithids=examp2,maxcp=4): 
load(file="/Users/sarahsullivan/Documents/SullivanDocs/Rwd/cutoff/ex2fitloops")
fitloopsobject<-ex2

bestfitsobject<-bestfits(fitres=fitloopsobject)
bicgraph(model=bestfitsobject)

fitpickobject<-fitpick(fitres=fitloopsobject,bestfits=bestfitsobject)

uncertobject<-uncert(fitpickobj=fitpickobject)
multcutobject<-multcut(uncertobj=uncertobject)

rawuncertgraph(uncertobj=uncertobject,multcutobj=multcutobject)
rawdistgraph(pickobj=fitpickobject)
rawhistcuts(uncertobj=uncertobject,multcutobj=multcutobject)

standindetobject<-standindet(uncertobj=uncertobject,multcutobj=multcutobject,cutcomp=1)
specindetobject<-specindet(uncertobj=uncertobject,multcutobj=multcutobject,cutcomp=1,certlevel=0.8)

cutuncertgraph(cutobj=standindetobject)
cutdistgraph(cutobj=standindetobject, pickobj=fitpickobject)
cutuncertgraph(cutobj=specindetobject)
cutdistgraph(cutobj=specindetobject, pickobj=fitpickobject)
outdataobject<-outputdata(standindet=standindetobject,specindet=specindetobject)
summaryobject<-summarytable(outdataobj=outdataobject, standindetobj=standindetobject, specindetobj=specindetobject)
outtable<-summaryobject$outtab
outtable

```{r while loop, echo=F}

stop<-F
iter<-0
comp<-0

  loopErr = function(data) {
    tryCatch(smsn.mix(data, nu =3, g = 5, get.init = TRUE, criteria = TRUE, group = TRUE, family = "Normal", calc.im=FALSE, obs.prob=TRUE, iter.max=5000, kmeans.param=list(iter.max = 100, n.start = 50)),
             error = function(e) {message(paste("Error in iteration", iter, sep=" ")); 
               NA}) 
  }

#fit5<-list()
#while (stop==F){
#  iter<-iter+1
#  set.seed(iter*10000+iter^2+iter)    
#  fit5[iter]<- loopErr(examp2$data)
#  if (is.na(fit5[iter])==F){
#    stop<-T
#  }
#}

>>>>

```{r loadfunctions, echo=F}

#-----------------------------------------------
#[IB] Load Packages
#-----------------------------------------------

library("mixsmsn")
library("sn")

#-----------------------------------------------
#[IC] Load Functions
#-----------------------------------------------

@@

#Runs loops of mixture model over different numbers of components and the normal and skew-normal distribution. This function accepts a data frame of the data and identifiers (variables must be entitled data and id respectively). This is the longest function to run and may fail, but can be re-run with adjustments using various options based on the error message output when the function fails. The default values are to do both skew-normal and normal approximation, with up to 5 components and 10 loops per combination of distribution and number of components. It is not reccommended to change these unless the default function fails.
#accepts a dataframe with columns id and data
fitloops<-function(datawithids,loops=10,maxcp=5){
  dists="Both"
  datawithids<-unique(datawithids)
  data<-datawithids$data
  datapts<-length(data)
  norms<-vector("list")
  errors<-matrix(0, nrow=maxcp, ncol=2)
  loopErr = function(data,i,j,family) {
    tryCatch(smsn.mix(data, nu =3, g = i, get.init = TRUE, criteria = TRUE, group = TRUE, family = family, calc.im=FALSE, obs.prob=TRUE, iter.max=5000, kmeans.param=list(iter.max = 50, n.start = 5)),
             error = function(e) {message(paste("Error in loop", j, "of", family, "with", i, "components", sep=" ")); 
               NA}) 
  }

@@

      norms$mc[[i]][[j]]<-vector("list", 15)
      norms$ms[[i]][[j]]<-vector("list", 15)
      if (dists=="Both") {
        norms$mc[[i]][[j]]<-loopErr(data,i,j,"Normal")  
        norms$ms[[i]][[j]]<-loopErr(data,i,j,"Skew.normal")
      } else if (dists=="Normal"){ 
        norms$mc[[i]][[j]]<-loopErr(data,i,j,"Normal")
      } else if (dists=="Skew.normal"){
        norms$ms[[i]][[j]]<-loopErr(data,i,j,"Skew.normal")
      } else {
        stop(print("dists must be specified as 'Both' or 'Normal' or 'Skew-normal'"))
      }

@@

    fitres<-vector("list")
    for (i in 1:maxcp){
      fitres$no[[i]]<-vector("list", 15)
      fitres$sn[[i]]<-vector("list", 15)
      fitres$no[[i]]<-norms$mc[[i]][[nobj$lmin$mc[i]]]
      fitres$sn[[i]]<-norms$ms[[i]][[nobj$lmin$ms[i]]]
    }
    fitres$loops<-loops
    fitres$maxcp<-maxcp
    fitres$datawithids<-datawithids
    fitres$errors<-errors
    return(fitres)
  } else {
    if (which.max(errors)<=maxcp){
      dist<-"Normal"
    } else {
      dist<-"Skew.normal"
    }
    if (ic<=2){
      stop(paste("More than two loops of the mixture modeling algorithm failed to converge for the", dist, "distribution with", ic, "component(s). First ensure that the necessary packages have been loaded correctly. If the packages are loaded correctly, check for extreme outliers (which may be dropped with extreme caution) or violations of skew-normal or normal distribution assumptions."))
    } else {
      stop(paste("More than two loops of the mixture modeling algorithm failed to converge for the", dist, "distribution with", ic, "component(s). Consider choosing a lower number of components (the default is 5) via the maxcp option of the fitloops function."))
    }
  }  
}

@@

  if (nobj$dist=="no"){
    nobj$distp="Normal"
    nobj$distf="Normal"
  } else if (nobj$dist=="sn") {
    nobj$distp="Skew-normal"
    nobj$distf="Skew-normal"
  } else {
    nobj$distp=""
  }

  model<-vector("list")
  model$sn2<-fitres$sn[[2]]
  model$snbest<-fitres$sn[[nobj$minind$sn[1]]]
  model$norm2<-fitres$no[[2]]
  model$normbest<-fitres$no[[nobj$minind$no[1]]]
  model$best<-nobj$best
  model$bictab<-data.frame(normal=c(NA)[rep(c(1), times=maxcp)], skew.normal=c(NA)[rep(c(1), times=maxcp)])
  model$bictab$normal<-nobj$no
  model$bictab$skew.normal<-nobj$sn
  componentnames<-vector()
  for (i in 1:maxcp){
    componentnames[i]<-paste(i, " Component(s):  ", sep="")  
  }    
  rownames(model$bictab)<-componentnames
  colnames(model$bictab)<-c("Normal", "Skew-normal")
  model$summary<-data.frame(desc=c(paste(nobj$distp, "with", nobj$minind[nobj$dist][1,1], "components ", sep=" "), paste("Skew-normal with", nobj$minind$sn, "components ", sep=" "), paste("Normal with", nobj$minind$no, "components ", sep=" "), "Skew-normal with 2 components ", "Normal with 2 components "), bic=c(nobj$best$bic, nobj$sn[nobj$minind$sn], nobj$no[nobj$minind$no], nobj$sn[2], nobj$no[2]))
  rownames(model$summary)<-c("Best Overall", "Best Skew-normal  ", "Best Normal", "Two Skew-normal", "Two Normal")
  colnames(model$summary)<-c("Description         ", "BIC  ")

@@

bicgraph<-function(model,title="BIC by type and number of components",setcolor=c("magenta", "blue", "black")){
  color<-setcolor
  buffer<-(max(model$bictab)-min(model$bictab))/20
  ylim<-c(min(model$bictab)-buffer, max(model$bictab)+buffer)
  if (is.na(model$bictab$Normal[1])==F) {
    plot(model$bictab$Normal, pch=1, col=color[1], main=title, xlab="Number of components", ylab="BIC", type="o",  xaxt="n", ylim=ylim)
    lines(model$bictab$`Skew-normal`, type="o", pch=2, lty=2, col=color[2])
  } else {
    plot(model$bictab$Normal, type="o", pch=2, lty=2, col=color[2], main=title, xlab="Number of components", ylab="BIC",  xaxt="n", ylim=ylim)
    lines(model$bictab$`Skew-normal`, pch=1, col=color[1], type="o")    
  }  
  points(model$bestdesc$ncomp,model$best$bic,type="o", pch="O", col=color[3],cex=2)
  axis(1, at = seq(1, length(model$bictab$Normal), by = 1))
  suppressWarnings(legend("topright", col=c(color[1], color[2], color[3]), lty=c(1,2,0), pch = c(1, 2, 1), legend=c("Normal", "Skew-normal", "Best by BIC"), cex=c(1,1,1), xjust=1, seg.len=3, title="Distribution"))

@@

  data=fitres$datawithids$data
  n=ncomp
  if (dist!="Normal" & dist!="Skew-normal") {
    stop("Distribution must be specified as either 'Normal' or 'Skew-normal'")
  } else if ((dist=="Normal" & is.na(fitres$no[[1]]$bic)==T)| (dist=="Skew-normal" & is.na(fitres$sn[[1]]$bic)==T) | ncomp>length(fitres$no) ){
    stop("Must choose a distribution and number of components that were fitted in the fitloops function")
  } else {
    if (dist=="Normal"){
      dist="no"
    } else {
      dist="sn"
    }

    if (dist=="no"){
      distp="Normal"
    } else if (dist=="sn") {
      distp="Skew-Normal"
    } else {
      distp=""
    }

@@

}
```

#Example 1
```{r example1, echo=T}

examp1<-read.csv(file="/Users/sarahsullivan/Documents/SullivanDocs/Rwd/cutoff/ex1.csv", colClasses=c("NULL", NA, NA))

hist(examp1$data, breaks=100, main="Histogram of Data")
dens1<-density(examp1$data)
plot(dens1, main="Density of Data")

#fitloopsobject<-fitloops(datawithids=examp2)
  #might not run due to 5 component solution not being easily optimized, if it doesn't run either load the full file, or run fitloopsobject<-fitloops(datawithids=examp2,maxcp=4): 
load(file="/Users/sarahsullivan/Documents/SullivanDocs/Rwd/cutoff/ex1fitloops")

fitloopsobject1<-ex1

bestfitsobject1<-bestfits(fitres=fitloopsobject1)
bicgraph(model=bestfitsobject1)

fitpickobject1<-fitpick(fitres=fitloopsobject1,bestfits=bestfitsobject1)

uncertobject1<-uncert(fitpickobj=fitpickobject1)
multcutobject1<-multcut(uncertobj=uncertobject1)

rawuncertgraph(uncertobj=uncertobject1,multcutobj=multcutobject1)
rawdistgraph(pickobj=fitpickobject1)
rawhistcuts(uncertobj=uncertobject1,multcutobj=multcutobject1)

standindetobject1<-standindet(uncertobj=uncertobject1,multcutobj=multcutobject1,cutcomp=1)
specindetobject1<-specindet(uncertobj=uncertobject1,multcutobj=multcutobject1,cutcomp=1,certlevel=0.8)

cutuncertgraph(cutobj=standindetobject1)
cutdistgraph(cutobj=standindetobject1, pickobj=fitpickobject1)
cutuncertgraph(cutobj=specindetobject1)
cutdistgraph(cutobj=specindetobject1, pickobj=fitpickobject1)
outdataobject1<-outputdata(standindet=standindetobject1,specindet=specindetobject1)
summaryobject1<-summarytable(outdataobj=outdataobject1, standindetobj=standindetobject1, specindetobj=specindetobject1)
outtable1<-summaryobject1$outtab
outtable1

@@

#Example 2

```{r example2, echo=T}

examp2<-read.csv(file="/Users/sarahsullivan/Documents/SullivanDocs/Rwd/cutoff/ex2.csv")

hist(examp2$data, breaks=100, main="Histogram of Data")
dens2<-density(examp2$data)
plot(dens2, main="Density of Data")

fitloopsobject2<-fitloops(datawithids=examp2)
bestfitsobject2<-bestfits(fitres=fitloopsobject2)
bicgraph(model=bestfitsobject2)

fitpickobject2<-fitpick(fitres=fitloopsobject2,bestfits=bestfitsobject2)

uncertobject2<-uncert(fitpickobj=fitpickobject2)
multcutobject2<-multcut(uncertobj=uncertobject2)

rawuncertgraph(uncertobj=uncertobject2,multcutobj=multcutobject2)
rawdistgraph(pickobj=fitpickobject2)
rawhistcuts(uncertobj=uncertobject2,multcutobj=multcutobject2)

standindetobject2<-standindet(uncertobj=uncertobject2,multcutobj=multcutobject2,cutcomp=1)
specindetobject2<-specindet(uncertobj=uncertobject2,multcutobj=multcutobject2,cutcomp=1,certlevel=0.8)

cutuncertgraph(cutobj=standindetobject2, xlim=c(min(examp2$data),3))
cutdistgraph(cutobj=standindetobject2, pickobj=fitpickobject2, xlim=c(min(examp2$data),3))
cutuncertgraph(cutobj=specindetobject2, xlim=c(min(examp2$data),3))
cutdistgraph(cutobj=specindetobject2, pickobj=fitpickobject2, xlim=c(min(examp2$data),3))
outdataobject2<-outputdata(standindet=standindetobject2,specindet=specindetobject2)
summaryobject2<-summarytable(outdataobj=outdataobject2, standindetobj=standindetobject2, specindetobj=specindetobject2)
outtable2<-summaryobject2$outtab
outtable2

@@
